---
title: "MATH601 Project 1"
author: "Eric O'Leary"
date: "December 5th, 2024"
output: pdf_document
---

# Creating Web-Based Graphics with Shiny

## What is Shiny?

Shiny is a package for R and Python created to help design interactive graphics for web applications. 
It is very simple to use, has robust documentation, tutorials, and example code, 
as well as plenty of features for customization.

```{r}
library(shiny)

runExample("01_hello")
```

Do note that the R console you use to run the app will be busy until you close the app,
which makes it more difficult to look at documentation or test code in R Studio.
I'd recommend designing all of your data wrangling + graphs without launching Shiny, 
then pasting the code into the app. 
This method also takes less time to run than launching Shiny multiple times.

The most intended way to run a Shiny project is to create a folder just for that project,
and to have a file named "app.R" with your Shiny code. You can then call runApp(directorypath)
in order to launch the app. Here, I have essentially written many app.R files which I run manually.


## A Pedagogical Example

To better illustrate how the package works and what it can do, let's create a simple example, step-by-step.

Firstly, we import the necessary libraries.

```{r}
library(shiny) # Package of focus
library(bslib) # Allows using Bootstrap/Sass (CSS framework) for page styling
library(ggplot2)
library(dplyr)
library(palmerpenguins) # Dataset of interest

penguins
```

Next, we create a blank page.

A Shiny app has two main features: a ui object, and a server function.
The ui object defines what the app looks like, and is very customizable, 
with different functions to create the object with different layouts 
as well as different UI elements to add to it.
The server function handles computations after the app has been launched in response to user input.

```{r}
# page_sidebar is easy format, but there are others
ui <- page_sidebar()

# Server must take input, output arguments, can optionally take a third
server <- function(input, output) {
  
}
```

```{r}
shinyApp(ui = ui, server = server)
```

Next, we add some basic elements to the ui.
IMPORTANT: UI elements are created by passing arguments to functions. 
As such, commas between these elements are necessary. 
Once elements start being nested, it is very easy to forget this.

```{r}
ui <- page_sidebar(
  # Adding text as the first item functions as a title
  h1("Relationships between measurements of Palmer penguins"),
  # Sidebar, good place to add controls for the main application
  sidebar = sidebar(
    # Text in sidebar
    "Text goes here",
    # Again, html format works so we can make our caption smaller
    p("Here's a caption", style="font-size:10px")
  ),
  # Card, most basic way to add sections to the main application
  card("Text")
)
```

```{r}
shinyApp(ui = ui, server = server)
```

Now, we want to add something interactive (even if it doesn't do anything yet).

```{r}
penguins

peng_numeric <- c("Bill Length" = "bill_length_mm",
                  "Bill Depth" = "bill_depth_mm",
                  "Flipper Length" = "flipper_length_mm",
                  "Body Mass" = "body_mass_g")


ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    
    selectInput(
      # Input id is internal label, important for later
      inputId = "x",
      # Label is what user sees above input method
      label = "Select a variable for x:",
      choices = peng_numeric
    ), # Don't forget comma
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  card("Text")
)
```

```{r}
shinyApp(ui, server)
```

Now that we have an option to select variables, we want to create a graph using them.
Since this now involves computation done after launching the app, we need to use server.

An important note on scoping: the server function has input and output as parameters, 
and can therefore access them like they exist. The UI object, however, does not have these
objects. So, in UI, we use strings, and in server, we use objects.

```{r}
# Input and output are objects that we read/write to modify the app
server <- function(input, output) {
  # To create dynamic output, we add it to the output object
  
  # Render plot is one of many render functions for different data types
  output$graph <- renderPlot({ # Main argument is an R expression to run
    # To access a selected variable, use input$(label)
    ggplot(penguins, aes(x = get(input$x), y = get(input$y))) +
      geom_point()
  })
}
```

This gives us a rendered plot of the variables we want, but does nothing to display it. Recall that the visual aspect of the app is controlled by ui:

```{r}
ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    selectInput(
      inputId = "x",
      label = "Select a variable for x:",
      choices = peng_numeric
    ),
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  # Since we used renderPlot, we need plotOutput
  # Since we saved the plot at output$graph, we use that label here
  card(plotOutput(outputId = "graph"))
)
```

```{r}
shinyApp(ui, server)
```

We now have an interactive graph mapping variables against each other. 
However, there are other variables in the dataset that may be of interest. 
So, let's add the ability to group by different variables. We'll also clean up the 
graph to make it look a bit more professional.

```{r}
peng_cats <- c("Species" = "species",
               "Island" = "island",
               "Sex" = "sex")

ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    selectInput(
      inputId = "x",
      label = "Select a variable for x:",
      choices = peng_numeric
    ),
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    # Add a choice of group
    # Using radio buttons for variety of not just using select
    radioButtons(
      inputId = "group",
      label = "Select a variable to color by:",
      choices = peng_cats
    ),
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  card(plotOutput(outputId = "graph"))
)



server <- function(input, output) {
  # Made prettier, and grouping by input exactly as expected
  output$graph <- renderPlot({
    ggplot(penguins, aes(x = get(input$x), y = get(input$y), 
                         color=get(input$group), shape=get(input$group))) +
      geom_point() +
      # Since names of our choices are prettier than the raw variables, 
      # we do a little transformation to retrieve them
      xlab(names(peng_numeric)[peng_numeric == input$x]) +
      ylab(names(peng_numeric)[peng_numeric == input$y]) +
      # Also, scale color and shape by selected categorical variable
      scale_color_discrete(names(peng_cats)[peng_cats == input$group]) +
      scale_shape_discrete(names(peng_cats)[peng_cats == input$group]) +
      labs(title = "Graph of different attributes of Palmer penguins")
  })
}

shinyApp(ui, server)
```

Now, we have a pretty graph with functioning interactivity! 
We could stop here, but let's add some linear regression to explain a few more concepts.

```{r}
ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    selectInput(
      inputId = "x",
      label = "Select a variable for x:",
      choices = peng_numeric
    ),
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    radioButtons(
      inputId = "group",
      label = "Select a variable to color by:",
      choices = peng_cats
    ),
    # We add a checkbox to ask whether to show regression or not
    checkboxInput(
      inputId = "regression",
      label = "Show linear regression?",
      value = FALSE
    ),
    # If the user wants regression, we then ask if they want it by group or not
    conditionalPanel(
      # The condition argument is a JavaScript expression, and can be very complex
      condition = "input.regression == true",
      checkboxInput(
        inputId = "reggroup",
        label = "Regress by group?",
        value = FALSE
      )
    ),
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  card(plotOutput(outputId = "graph"))
)
```

So, we know that we want the option to add a regression line, and we also want to 
be able to do so with or without following the user's selected group.
For this, we will use reactivity.

Reactivity is a very important concept when making larger applications, using larger data,
or handling data from the web such as with live updates.
To explain, we want to look at the structure of the application and talk about code efficiency.

When writing an app.R file, it will look something like this:

```{r, eval=FALSE}
# Code here will run once, when the app is launched from the creator
# This should include loading libraries, tidying data, and defining variables
# UI is a variable, so it will be run once at this stage as well

library(shiny)
library(ggplot2)
library(palmerpenguins)

peng_cats <- c("Species" = "species",
               "Island" = "island",
               "Sex" = "sex")

peng_numeric <- c("Bill Length" = "bill_length_mm",
                  "Bill Depth" = "bill_depth_mm",
                  "Flipper Length" = "flipper_length_mm",
                  "Body Mass" = "body_mass_g")

ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    selectInput(
      inputId = "x",
      label = "Select a variable for x:",
      choices = peng_numeric
    ),
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    radioButtons(
      inputId = "group",
      label = "Select a variable to color by:",
      choices = peng_cats
    ),
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  card(plotOutput(outputId = "graph"))
)

server <- function(input, output) {
  # The second place code is run is inside the server function
  # This code is run once PER USER when they first connect to the application
  # Anything computation that can be taken out of here should be for efficiency
  # Render functions will also be run once when the user connects to the app
  
  output$graph <- renderPlot({
    # Any time a variable used inside a render expression is updated by the user,
    # the ENTIRE render expression will be re-run
    
    # If our render function scrapes the most recent version of a website's data,
    # our application will be using an impolite amount of bandwidth
    ggplot(penguins, aes(x = get(input$x), y = get(input$y), 
                         color=get(input$group), shape=get(input$group))) +
      geom_point() +
      xlab(names(peng_numeric)[peng_numeric == input$x]) +
      ylab(names(peng_numeric)[peng_numeric == input$y]) +
      scale_color_discrete(names(peng_cats)[peng_cats == input$group]) +
      scale_shape_discrete(names(peng_cats)[peng_cats == input$group]) +
      labs(title = "Graph of different attributes of Palmer penguins")
  })
}

shinyApp(ui, server)
```

If we assume our application will have thousands of concurrent users, we want to
minimize the amount of computation done per click.

Reactivity has the program track whether a certain expression has been updated and
recalculate only if necessary. 

One example is if we wanted to filter our data and perform a transformation after.
Suppose we have a dataset with time and income as variables. If the user is only interested
in a specific time frame, then we only want to use what they select because it will
be faster to transform than the entire dataset. If they then choose to scale income 
logarithmically , we don't want to transform the dataset or recalculate the entire graph,
just modify what we already have. 

```{r, eval = FALSE}
# This is a completely fictitious example, purely for demonstration
server <- function(input, output) {
  
  # Since this only depends on the user's input for time,
  # it will only recalculate if those inputs change
  selectedData <- reactive({
    data |> 
      filter(time >= input$timelower & time <= input$timeupper)
  })
  
  # If the user changes their mind about logarithmic scaling, then
  # this will recalculate. But, if they don't change the time range, the data 
  # will not need to be recalculated
  incomePlot <- reactive({
    ggplot(selectedData(), aes(x=time, y=income)) +
      (if(input$logscale == TRUE) {
        scale_y_continuous(trans="log")
      })
  })
  
  output$plot <- renderPlot({
    incomePlot()
  })
}
```

Now, we apply those principles to our penguins example.

```{r}
ui <- page_sidebar(
  h1("Relationships between Measurements of Penguins"),
  sidebar = sidebar(
    "Choose variables and groupings to investigate relationships between different variables",
    selectInput(
      inputId = "x",
      label = "Select a variable for x:",
      choices = peng_numeric
    ),
    selectInput(
      inputId = "y",
      label = "Select a variable for y:",
      choices = peng_numeric
    ),
    radioButtons(
      inputId = "group",
      label = "Select a variable to color by:",
      choices = peng_cats
    ),
    checkboxInput(
      inputId = "regression",
      label = "Show linear regression?",
      value = FALSE
    ),
    conditionalPanel(
      condition = "input.regression == true",
      checkboxInput(
        inputId = "reggroup",
        label = "Regress by group?",
        value = FALSE
      )
    ),
    p("Data from palmerpenguins package", style="font-size:10px")
  ),
  card(plotOutput(outputId = "graph"))
)

server <- function(input, output) {
  aesReg <- reactive({
    if(input$reggroup) {
      aes(group = get(input$group), color = get(input$group))
    } else {
      aes()
    }
  })
  
  pengPlot <- reactive({
    # Slightly changed; moved grouping to geom_point so geom_smooth can choose
    ggplot(penguins, aes(x = get(input$x), y = get(input$y))) +
      geom_point(aes(color=get(input$group), shape=get(input$group))) +
      xlab(names(peng_numeric)[peng_numeric == input$x]) +
      ylab(names(peng_numeric)[peng_numeric == input$y]) +
      scale_color_discrete(names(peng_cats)[peng_cats == input$group]) +
      scale_shape_discrete(names(peng_cats)[peng_cats == input$group]) +
      labs(title = "Graph of different attributes of Palmer penguins") +
      (if(input$regression == TRUE) {
        geom_smooth(method="lm", aesReg())
      })
  })
  
  output$graph <- renderPlot({
    pengPlot()
  })
}
```

```{r}
shinyApp(ui, server)
```


## Deploying your Project

Once you have a completed project, an easy way to post it for others to see is shinyapps.io.
This service allows users to host Shiny applications online with a free option.
It is very easy to setup and use, and then 

For an example, I have a full project with some more complex uses of Shiny
available for viewing at https://eoleary.shinyapps.io/project1/.


## Sources

In general, check out the very robust documentation at https://shiny.posit.co. 
For beginner tutorials, check out https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/. 
For using ShinyApps, the website itself, https://www.shinyapps.io/, has an easy tutorial.