---
  title: "SoA Case Study 2025 Work"
  author: "Eric O'Leary"
---

This is a workbook for the 2025 SoA case study for team AES Consulting.
This is largely unorganized scratch work, and is not the complete model used in the final version.
However, this does represent the team's data exploration and preliminary to intermediate analysis, primarily organized and conducted by me.



Team AES Consulting


Notes:
Navaldia only region with undetermined hazard level

Hazard:
High hazard - high property damage and casualties
Significant hazard - high property damage

Assessment:
Satisfactory - fine
Fair - no safety deficiency under normal conditions
Poor - normal realistic conditions could fail
Unsatisfactory - emergency action required


# Importing Libraries/Data

```{r}
library(readxl)
library(tidyverse)
library(mice)
```

```{r}
damraw <- read.csv("srcsc-2025-dam-data-for-students.csv")
econdataraw <- read_excel("srcsc-2025-economic-data-summary.xlsx", skip = 10)
```

# Analysis

## Inflation and Interest Rate

```{r}
econdata <- econdataraw
colnames(econdata) <- c("Year", "Inflation", "Overnight", "Spot1Yr", "Spot10Yr")
  

# https://data.bls.gov/pdq/SurveyOutputServlet
usinflation <- read_xlsx("USInflation.xlsx", skip=11)
usyearly <- usinflation |>
  dplyr::select(-c("HALF1", "HALF2")) |> 
  pivot_longer(
    names_to = "Month",
    cols = -c("Year"),
    values_to = "Inflation"
  ) |> 
  group_by(Year) |>
  summarize(us = (prod(1+Inflation/100) ^ (1/12))-1)

both_inflation <- merge(econdata |> select(Year, Inflation), usyearly, by="Year")

both_inflation |> 
  rename(tarrodan = Inflation) |>
  pivot_longer(
    cols = c("tarrodan", "us"),
    values_to = "inflation",
    names_to = "country"
  ) |> 
  ggplot(aes(x=Year, y=inflation, color=country)) + 
    geom_line(lwd=.5) +
    scale_color_discrete(name="Country", labels=c("Tarrodan", "United States")) +
    labs(title = "Yearly Inflation, United States and Tarrodan") + 
    ylab("Inflation")
ggsave("InflationUSTarrodan.png",
       width=1920,
       height=1080,
       units="px")

diff_inflation <- both_inflation |>
  mutate(diff = Inflation - us)
```

```{r}
yravg = c(NA)
for(i in seq(1:(nrow(econdata)-9))) {
  sums = c(NA)
  for(j in 0:9) {
    sums[j] = 1 + econdata$Spot1Yr[[i + j]]
  }
  yravg[i] = prod(sums) ^ (1/10)
}
min(yravg)
mean(yravg)
max(yravg)

yravg = c(NA)
for(i in seq(1:(nrow(usyearly)-9))) {
  sums = c(NA)
  for(j in 0:9) {
    sums[j] = 1 + usyearly$us[[i + j]]
  }
  yravg[i] = prod(sums) ^ (1/10)
}
mean(yravg)
```


```{r}
# Just look at error term scatter plot
ggplot(diff_inflation, aes(x=Year, y=diff)) + geom_point()


# 480x270 for slides, but 960x540 is better for larger scale
png("InflationHistogramUSTarrodan.png", width=960, height=540, units="px")
hist(diff_inflation$diff, breaks=20, main="Histogram of Difference in Inflation Rates, US and Tarrodan", xlab="Difference in inflation rate")
dev.off()

png("InflationDensityUSTarrodan.png", width=960, height=540, units="px")
plot(density(diff_inflation$diff), main="Density of Difference in Inflation Rates, US and Tarrodan", xlab="Difference in inflation rates")
dev.off()

png("InflationQQNormUSTarrodan.png", width=960, height=540, units="px")
qqnorm(diff_inflation$diff)
dev.off()

with(diff_inflation, ks.test(diff, y="pnorm", mean=mean(diff), sd=sd(diff)))
```

```{r}
usovernightraw <- read_xlsx("USOvernight.xlsx")
usovernight <- usovernightraw |> 
  select(c(1,2,3))
colnames(usovernight) <- c("Date", "Type", "Rate")
usovernight <- usovernight |> 
  filter(Type == "OBFR") |> 
  mutate(Date = as.Date(Date, "%m/%d/%Y")) |>
  mutate(Year = year(Date)) |>
  group_by(Year) |>
  summarize(Rate = mean(Rate) / 100) # Is mean best way to summarize

both_overnight <- merge(econdata |> select(Year, Overnight), usovernight, by="Year")
ggplot(both_overnight |>
         rename(Tarrodan = Overnight,
                US = Rate) |>
         pivot_longer(
           values_to = "rate",
           names_to = "country",
           cols = c(2,3)
         ), aes(x=Year, y=rate, color=country)) + 
  geom_line() + 
  labs(title="Overnight Rate, US and Tarrodan")
# Due to small amount of data, just say they're close enough
ggsave("OvernightRates.png", width=1920, height=1080, units="px")

us1yrraw <- read.csv("US1YrSpot.csv")
us1yr <- us1yrraw |> 
  mutate(Date = as.Date(observation_date, "%Y-%m-%d")) |>
  mutate(Year = year(Date)) |>
  group_by(Year) |>
  summarize(Rate = mean(DGS1, na.rm=T) / 100)

both_1yr <- merge(econdata |> select(Year, Spot1Yr), us1yr, by="Year")
ggplot(both_1yr |>
         rename(Tarrodan = Spot1Yr,
                US = Rate) |>
         pivot_longer(
           values_to = "rate",
           names_to = "country",
           cols = c(2,3)
         ), aes(x=Year, y=rate, color=country)) + 
  geom_line() + 
  labs(title="1-Year Spot Rate, US and Tarrodan")
ggsave("OneYearSpot.png", width=1920, height=1080, units="px")

diff_1yr <- both_1yr |> mutate(diff = Spot1Yr - Rate)
hist(diff_1yr$diff, breaks=20, main="Histogram of Difference in 1-Year Spot Rates, US and Tarrodan", xlab="Difference in spot rates")

plot(density(diff_1yr$diff), main="Density of Difference in 1-Year Spot Rates, US and Tarrodan", xlab="Difference in spot rates")

qqnorm(diff_1yr$diff)
qqnorm((diff_1yr |> filter(Year != 1987))$diff)
# In this case, rather than random noise, it's just so close in almost every case
# that we just say they're practically the same

  
us10yrraw <- read.csv("US10YrSpot.csv")
us10yr <- us10yrraw |> 
  mutate(Date = as.Date(observation_date, "%Y-%m-%d")) |>
  mutate(Year = year(Date)) |>
  group_by(Year) |>
  summarize(Rate = mean(DGS10, na.rm=T) / 100)

both_10yr <- merge(econdata |> select(Year, Spot10Yr), us10yr, by="Year")
ggplot(both_10yr |>
         rename(Tarrodan = Spot10Yr,
                US = Rate) |>
         pivot_longer(
           values_to = "rate",
           names_to = "country",
           cols = c(2,3)
         ), aes(x=Year, y=rate, color=country)) + 
  geom_line() + 
  labs(title="10-Year Spot Rate, US and Tarrodan")
ggsave("TenYearSpot.png", width=1920, height=1080, units="px")

diff_10yr <- both_10yr |> mutate(diff = Spot10Yr - Rate)
hist(diff_10yr$diff, breaks=20, main="Histogram of Difference in 10-Year Spot Rates, US and Tarrodan", xlab="Difference in spot rates")

plot(density(diff_10yr$diff), main="Density of Difference in 10-Year Spot Rates, US and Tarrodan", xlab="Difference in spot rates")

qqnorm(diff_10yr$diff)
qqnorm((diff_10yr |> filter(Year != 1987))$diff)
```



# Data Cleaning

```{r}
damdata <- damraw

# Human + machine readable column names
colnames(damdata) <- c("ID", "Region", "Regulated", "Purpose", "Type", "Height", "Length", "Volume", "YearCompleted", "YearsModified", "Surface", "Drainage", "Spillway", "LastInspection", "InspectionFreq", "DistanceToCity", "Hazard", "Assessment", "AssessmentDate", "ProbFailure", "LGFProperty", "LGFLiability", "LGFBusiness")

# Replace empty string with NA
damdata <- replace(damdata, damdata == "", NA)

# Modify column types for ease of computing
damdata <- damdata |> 
  mutate(
    Region = as.factor(Region),
    Regulated = as.factor(Regulated),
    Purpose = as.factor(Purpose),
    Type = as.factor(Type),
    Spillway = as.factor(Spillway),
    LastInspection = as.Date(LastInspection, format="%d/%m/%Y"),
    Hazard = factor(Hazard, levels=c("Low", "Significant", "High", "Undetermined")),
    # UNKNOWN IF THIS PROPER RANKING
    Assessment = factor(Assessment, levels=c("Unsatisfactory", "Poor", "Fair", "Satisfactory", "Not Rated", "Not Available")),
    AssessmentDate = as.Date(AssessmentDate, "%d/%m/%Y")
  ) |> # Set height 0 to missing (since height 0 impossible)
  mutate(Height = na_if(Height, 0)) |>
  # LastModified -> max(YearsModified, YearCompleted)
  mutate(YearsModified = pmax(YearCompleted,
                             as.numeric(gsub("([0-9]{4}.?)?([0-9]+).*$",
                                             "\\2", damdata$YearsModified)),
                             na.rm=T)) |>
  rename(LastModified = YearsModified)

# An inspection exists with year 2031; assume mistake and set to missing
damdata$LastInspection[which(year(damdata$LastInspection) == 2031)] <- as.Date(NA)
```

# Data Exploration

```{r}
# Getting minimum total loss; rough estimate of yearly loss
damdata |> mutate(T = rowSums(damdata[21:23], na.rm=T)) |> arrange(T)
```

```{r}
# List number of NA's in each column
sapply(damdata, function(y) sum(length(which(is.na(y)))))

# Proportion of NA's per column
sapply(damdata, function(y) sum(length(which(is.na(y))))) / nrow(damdata) * 100
```

```{r}
# Expected loss (assuming NA = 0)
damdata |> 
  mutate(TotalLoss = rowSums(damdata[21:23], na.rm=T)) |>
  mutate(ExpectedLoss = ProbFailure * TotalLoss)  |> 
  group_by(Region) |>
  summarize(Count = n(),
            RegionalLoss = sum(ExpectedLoss),
            PropLoss = RegionalLoss / Count)

damdata |> 
  group_by(Region) |> 
  summarize(ExpectedFailureCount = sum(ProbFailure))


damdata |> group_by(Region) |> summarize(n())

damdata |> 
  mutate(TotalLoss = rowSums(damdata[21:23], na.rm=T)) |>
  mutate(ExpectedLoss = ProbFailure * TotalLoss) |> 
  summarize(MinExpectedLoss = sum(ExpectedLoss))

# Data is in millions

# 775bil/10yr, say 80bil/yr, 
```

```{r}
# Yearly calculation
# Assuming exponential distribution, we are given F(10) and want lambda
# Or, F(10) = 1-e^{-lambda 10}
# So, e^{-lambda 10} = 1-F(10)
# => -lambda 10 = log(1-F(10))
# => lambda = -log(1-F(10)) / 10
# ALSO, loss is TotalLoss * R.V. ~ Exp(HazardRate), so E(loss) = TotalLoss * RV
# Similarly, Var(loss) = Var(TotalLoss*RV) = TotalLoss^2 * RV
# And, variance of sum is sum of variance for indep, so sum is total variance
damdata |> 
  mutate(HazardRate = -log(1-ProbFailure) / 10,
         TotalLoss = rowSums(damdata[21:23], na.rm=T),
         ExpectedLoss = TotalLoss * ProbFailure,
         ExpectedLossOneYr = TotalLoss * (1-exp(-HazardRate)),
         Variance = TotalLoss^2 * ProbFailure,
         VarianceOneYr = TotalLoss^2 * (1-exp(-HazardRate))) |>
  summarize(TenYrLoss = sum(ExpectedLoss),
            OneYrLoss = sum(ExpectedLossOneYr),
            TenYrSD = sqrt(sum(Variance)),
            OneYrSD = sqrt(sum(VarianceOneYr))) |>
  summarize(OneYrLoss * 1000000 / 94699280,
            TenYrLoss * 1000000 / 94699280)
```

# Data Imputation

```{r}
# Imputing important missing data only
# Height, assessments of any missing kind, LGF

# Standardization to avoid unsolvable equations
tempdam <- damdata |> mutate(across(where(is.numeric),  ~(scale(.) %>% as.vector)))

damstats <- damdata |> 
  summarize_if(is.numeric, list(mean = mean, sd = sd), na.rm=T)

damimpprep <- tempdam |> 
  mutate(Assessment = ifelse(Assessment %in% c("Unsatisfactory", "Poor", "Fair",
                                               "Satisfactory"), 
                             Assessment, 
                             NA)) |>
  mutate(Assessment = factor(Assessment, 
                             levels = c(1,2,3,4),
                             labels=c("Unsatisfactory", "Poor", "Fair",
                                      "Satisfactory"))) |>
  select(-c("LastInspection", "AssessmentDate"))

# No colinear columns, which is desirable
mice:::find.collinear(damimpprep)

# Quickpred for faster calculation
init <- mice(damimpprep, pred=quickpred(damimpprep), maxit=0)

pred <- init$predictorMatrix
# Do not impute on columns of no interest
pred[1:5, ] = 0
pred[7:16, ] = 0

impute_cols <- mice(damimpprep, pred=pred, m=3, maxit=30)

# saveRDS(impute_cols, file="imputedObject.RDS")


# png("ImputationTraceplots.png", width=960, height=4140, units="px")
plot(impute_cols, layout=c(2,21))
# dev.off()

# Doesn't work
# plot(impute_cols, ~ Height + Assessment + LGFProperty + LGFLiability + LGFBusiness)

# densityplot(impute_cols, ~ Height + Assessment + LGFProperty + LGFLiability + LGFBusiness)
```

```{r}
impute_cols <- readRDS("imputedObject.RDS")
```

```{r}
# Un-standardize imputed data
# NOTE: Does not quality check of unreasonable values (such as negatives)
impeddam <- complete(impute_cols) |> 
  mutate(Height = (Height * damstats[1, "Height_sd"]) + damstats[1, "Height_mean"], 
         Length = (Length * damstats[1, "Length_sd"]) + damstats[1, "Length_mean"],
         Volume = (Volume * damstats[1, "Volume_sd"]) + damstats[1, "Volume_mean"],
         YearCompleted = (YearCompleted * damstats[1, "YearCompleted_sd"]) + damstats[1, "YearCompleted_mean"],
         LastModified = (LastModified * damstats[1, "LastModified_sd"]) + damstats[1, "LastModified_mean"],
         Surface = (Surface * damstats[1, "Surface_sd"]) + damstats[1, "Surface_mean"],
         Drainage = (Drainage * damstats[1, "Drainage_sd"]) + damstats[1, "Drainage_mean"],
         InspectionFreq = (InspectionFreq * damstats[1, "InspectionFreq_sd"]) + damstats[1, "InspectionFreq_mean"],
         DistanceToCity = (DistanceToCity * damstats[1, "DistanceToCity_sd"]) + damstats[1, "DistanceToCity_mean"],
         LGFProperty = (LGFProperty * damstats[1, "LGFProperty_sd"]) + damstats[1, "LGFProperty_mean"],
         LGFLiability = (LGFLiability * damstats[1, "LGFLiability_sd"]) + damstats[1, "LGFLiability_mean"],
         LGFBusiness = (LGFBusiness * damstats[1, "LGFBusiness_sd"]) + damstats[1, "LGFBusiness_mean"],
         ProbFailure = (ProbFailure * damstats[1, "ProbFailure_sd"]) + damstats[1, "ProbFailure_mean"])

# write.csv(impeddam, "ImputedDamData.csv", row.names=FALSE)
# impeddam <- read.csv("ImputedDamData.csv")
```



```{r}
height_bins_feet <- c(15, 25, 50, 100, 200)
hbm <- height_bins_feet / 3.28084 # Convert to meters

# Price bins grouping by height bin within repair type bin
price_bins_heightsmaller <- c(400, 790, 1410, 1360, 3080, 9180,
                              1380, 1890, 4000, 4800, 20000, 26340,
                              2870, 2670, 6230, 8580, 23840, 95300) / (1000*1.048)
# Converted to Qalkoons at 1.048USD to 1.000Q
```


```{r}
imped_damcosts <- impeddam |> 
  mutate(Age = 2025-YearCompleted) |> 
  mutate(TotalLoss = rowSums(impeddam[19:21], na.rm=T)) |>
  select(c("Region", "Height", "Age", "Hazard", "Assessment", "ProbFailure", "TotalLoss")) |> 
  mutate(ExpectedLoss = ProbFailure * TotalLoss) |>
  mutate(Lambda = -log(1-ProbFailure) / 10) |>
  mutate(ExpectedLoss1yr = TotalLoss * (1-exp(-Lambda))) |>
  rename(NewAssessment = Assessment) |> # To fit other code easier
  mutate(HeightBin = case_when(Height<=hbm[1] ~ 1,
                               Height>hbm[1] & Height<=hbm[2] ~ 2,
                               Height>hbm[2] & Height<=hbm[3] ~ 3,
                               Height>hbm[3] & Height<=hbm[4] ~ 4,
                               Height>hbm[4] & Height<=hbm[5] ~ 5,
                               Height>hbm[5] ~ 6)) |> 
  mutate(RepairBin = case_when(Age<50 & NewAssessment != "Satisfactory" ~ 1,
                               Age>=50 & NewAssessment == "Fair" ~ 2,
                               Age>=50 & (NewAssessment == "Poor" | NewAssessment == "Unsatisfactory") ~ 3, .default = 0)) |> 
  mutate(ExpectedRepair = case_when(RepairBin == 1 ~
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[1],
                                        HeightBin == 2 ~ price_bins_heightsmaller[2],
                                        HeightBin == 3 ~ price_bins_heightsmaller[3],
                                        HeightBin == 4 ~ price_bins_heightsmaller[4],
                                        HeightBin == 5 ~ price_bins_heightsmaller[5],
                                        HeightBin == 6 ~ price_bins_heightsmaller[6]
                                      ),
                                    RepairBin == 2 ~ 
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[7],
                                        HeightBin == 2 ~ price_bins_heightsmaller[8],
                                        HeightBin == 3 ~ price_bins_heightsmaller[9],
                                        HeightBin == 4 ~ price_bins_heightsmaller[10],
                                        HeightBin == 5 ~ price_bins_heightsmaller[11],
                                        HeightBin == 6 ~ price_bins_heightsmaller[12]
                                      ),
                                    RepairBin == 3 ~ 
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[13],
                                        HeightBin == 2 ~ price_bins_heightsmaller[14],
                                        HeightBin == 3 ~ price_bins_heightsmaller[15],
                                        HeightBin == 4 ~ price_bins_heightsmaller[16],
                                        HeightBin == 5 ~ price_bins_heightsmaller[17],
                                        HeightBin == 6 ~ price_bins_heightsmaller[18]
                                      ))
         )

damcosts <- imped_damcosts

```


```{r}
imped_damcosts |> 
  group_by(Region) |> 
  summarize(T = sum(ExpectedLoss1yr)) |> 
  mutate(T = 1000000 * T / c(15911120, 2967382, 16323908))

imped_damcosts |> 
  summarize(T = sum(ExpectedLoss1yr) * 1000000 / 35202410)

# Year 1 premiums  divided per dam
```


```{r non-imputed}
# Damcosts for non-imputed data
assessments <- factor(c("Unsatisfactory", "Poor", "Fair", "Satisfactory"),
                      levels=c("Unsatisfactory", "Poor", "Fair", "Satisfactory"))
aprops <- (damdata |>
  filter(!(is.na(Assessment) | Assessment == "Not Rated" | Assessment == "Not Available")) |>
  group_by(Region, Assessment) |>
  summarize(Assessment_Prop = n()) |>
  mutate(Assessment_Prop = Assessment_Prop / sum(Assessment_Prop)))$Assessment_Prop

set.seed(123)
damcosts <- damdata |> 
  mutate(Age = 2025-YearCompleted) |> 
  mutate(TotalLoss = rowSums(damdata[21:23], na.rm=T)) |>
  select(c("Region", "Height", "Age", "Hazard", "Assessment", "ProbFailure", "TotalLoss")) |> 
  mutate(ExpectedLoss = ProbFailure * TotalLoss) |>
  mutate(Lambda = -log(1-ProbFailure) / 10) |>
  mutate(ExpectedLoss1yr = TotalLoss * (1-exp(-Lambda))) |>
  # Assumptions of original pricing method
  mutate(Age = replace_na(Age, 51)) |> # If age missing, assume older than 50
  # Replace missing assessment by sampling from regional proportions
  mutate(NewAssessment = 
           ifelse(is.na(Assessment) | 
                    Assessment == "Not Rated" | Assessment == "Not Available",
                  case_when(Region == "Flumevale" ~ 
                              sample(assessments, 1, replace=T, prob=aprops[1:4]),
                            Region == "Lyndrassia" ~ 
                              sample(assessments, 1, replace=T, prob=aprops[5:8]),
                            Region == "Navaldia" ~ 
                              sample(assessments, 1, replace=T, prob=aprops[9:12])),
         as.factor(Assessment))) |>
  mutate(NewAssessment = factor(NewAssessment, levels=c(1,2,3,4), labels=c("Unsatisfactory", "Poor", "Fair", "Satisfactory"))) |>
  mutate(HeightBin = case_when(is.na(Height) ~ 0,
                               Height<=hbm[1] ~ 1,
                               Height>hbm[1] & Height<=hbm[2] ~ 2,
                               Height>hbm[2] & Height<=hbm[3] ~ 3,
                               Height>hbm[3] & Height<=hbm[4] ~ 4,
                               Height>hbm[4] & Height<=hbm[5] ~ 5,
                               Height>hbm[5] ~ 6)) |> 
  mutate(RepairBin = case_when(Age<50 & NewAssessment != "Satisfactory" ~ 1,
                               Age>=50 & NewAssessment == "Fair" ~ 2,
                               Age>=50 & (NewAssessment == "Poor" | NewAssessment == "Unsatisfactory") ~ 3, .default = 0)) |> 
  mutate(ExpectedRepair = case_when(RepairBin == 1 ~
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[1],
                                        HeightBin == 2 ~ price_bins_heightsmaller[2],
                                        HeightBin == 3 ~ price_bins_heightsmaller[3],
                                        HeightBin == 4 ~ price_bins_heightsmaller[4],
                                        HeightBin == 5 ~ price_bins_heightsmaller[5],
                                        HeightBin == 6 ~ price_bins_heightsmaller[6]
                                      ),
                                    RepairBin == 2 ~ 
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[7],
                                        HeightBin == 2 ~ price_bins_heightsmaller[8],
                                        HeightBin == 3 ~ price_bins_heightsmaller[9],
                                        HeightBin == 4 ~ price_bins_heightsmaller[10],
                                        HeightBin == 5 ~ price_bins_heightsmaller[11],
                                        HeightBin == 6 ~ price_bins_heightsmaller[12]
                                      ),
                                    RepairBin == 3 ~ 
                                      case_when(
                                        HeightBin == 1 ~ price_bins_heightsmaller[13],
                                        HeightBin == 2 ~ price_bins_heightsmaller[14],
                                        HeightBin == 3 ~ price_bins_heightsmaller[15],
                                        HeightBin == 4 ~ price_bins_heightsmaller[16],
                                        HeightBin == 5 ~ price_bins_heightsmaller[17],
                                        HeightBin == 6 ~ price_bins_heightsmaller[18]
                                      ))
         )

# Note: This has a few entries with missing heights, 
# which cannot be given repair costs by the above methodology
damdata |> filter(is.na(Height))
```

# Modeling

```{r}
# Expected total cost of repairs for all dams
damcosts |>
  filter(HeightBin != 0) |>
  filter(RepairBin != 0) |> # Disregard dams with no need for repairs
  select(-c("RepairBin", "HeightBin")) |>
  summarize(Rep = sum(ExpectedRepair))

# Calculate average probability of failure for a "satisfactory" dam
probs_by_region <- (damdata |> 
  filter(Assessment == "Satisfactory") |>
  group_by(Region) |>
  summarize(Probs = mean(ProbFailure)))$Probs

# Calculate expected repair costs + savings for all profitable dams
damcosts |>
  mutate(NewProb = case_when(Region == "Flumevale" ~ probs_by_region[1],
                             Region == "Lyndrassia" ~ probs_by_region[2],
                             Region == "Navaldia" ~ probs_by_region[3])) |>
  mutate(ExpectedNewLoss = NewProb * TotalLoss) |>
  mutate(ExpectedSavings = ExpectedLoss - (ExpectedRepair + ExpectedNewLoss)) |>
  filter(ExpectedSavings >= 0) |> # Comment out line for all dams
  summarize(sum(ExpectedRepair), sum(ExpectedSavings))
```


```{r}
damexps <- damcosts |> 
  # mutate(NeedsRep = ifelse(RepairBin != 0 | Hazard == "High" | 
                             # Hazard == "Significant" | ExpectedSavings >= 0, 1, 0)) |>
  # filter(HeightBin != 0) |> # Only needed for non-imputed
  # filter(RepairBin != 0) |> # Disregard dams with no need for repairs
  select(-c("RepairBin", "HeightBin", "Height", "Age")) |>
  mutate(NewProb = case_when(Region == "Flumevale" ~ probs_by_region[1],
                             Region == "Lyndrassia" ~ probs_by_region[2],
                             Region == "Navaldia" ~ probs_by_region[3])) |>
  mutate(NewLambda = -log(1-NewProb) / 10) |>
  mutate(NewProbYr = 1-exp(-NewLambda)) |>
  mutate(ExpectedNewLoss = NewProb * TotalLoss) |>
  mutate(ExpectedSavings = ExpectedLoss - (ExpectedRepair + ExpectedNewLoss)) |>
  # Find dams where repair is profitable or hazard is high/significant, 
  # flag expected savings 
  mutate(ExpectedSavings = ifelse(
    ExpectedSavings >= 0 | Hazard == "High" | Hazard == "Significant",
    ExpectedSavings, NA)
  ) |>
  # Adding significant as well only adds ~50 dams of 7.7k (unimputed)
  # filter(ExpectedSavings >= 0 | Hazard == "High" | Hazard == "Significant") |>
  mutate(ExpectedNewLossYrly = TotalLoss * NewProbYr) |>
  # THIS EXPECTED SAVINGS DOES NOT COUNT FOR REPAIR COSTS
  mutate(ExpectedSavingsYrly = ExpectedLoss1yr - ExpectedNewLossYrly)#  |>
  arrange(desc(ExpectedSavings))
```


```{r}
# TODO: Organize code to be easily reproducible
# Define some functions for sensitivity, like the whole model
# as a function taking climate change as a parameter,
# maybe dataset too so we can calc repairs only and full
# Then save the resulting dataframe as variable, apply to summary/graphing below
# Also apply inflation/interest parameters to those things

model_dams <- function(damexps, clim_coef) {
  saving_splits <- quantile(x=(damexps |> 
    filter(ExpectedSavings >= 0 | Hazard == "High" | Hazard == "Significant"))$ExpectedSavings,
    c(.9,.8,.7,.6,.5,.4,.3,.2,.1,0),
    na.rm=T
  )
# UNADJUSTED FOR INFLATION

# If repaired, set lambda to minimum of region average and its baseline, 
# unadjusted for potential climate change
# This assumption may be flawed, but it's what we're going with
# If not repaired this year, use previous year's
# In case satisfactory dams increase from climate change, this is needed
  damexps |>
    mutate(RepYr1 = ifelse(ExpectedSavings >= saving_splits[[1]], 1, 0)) |>
    mutate(AsmntYr1 = if_else(RepYr1 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr1 = if_else(RepYr1 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr1 = case_when(
      AsmntYr1 == "Satisfactory" ~ LambdaYr1 * clim_coef[1],
      AsmntYr1 == "Fair" ~ LambdaYr1 * clim_coef[2],
      AsmntYr1 == "Poor" ~ LambdaYr1 * clim_coef[3],
      AsmntYr1 == "Unsatisfactory" ~ LambdaYr1 * clim_coef[4]
      )) |>
    mutate(ProbYr1 = 1-exp(-LambdaYr1)) |>
    mutate(ExpectedLossYr1 = TotalLoss * ProbYr1) |>
    mutate(RepCostYr1 = ifelse(RepYr1 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr1 = ifelse(RepYr1 == 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr2 = ifelse(ExpectedSavings >= saving_splits[[2]], RepYr1 + 1, 0)) |>
    mutate(AsmntYr2 = if_else(RepYr2 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr2 = if_else(RepYr2 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr2 = case_when(
      AsmntYr2 == "Satisfactory" ~ LambdaYr2 * clim_coef[1],
      AsmntYr2 == "Fair" ~ LambdaYr2 * clim_coef[2],
      AsmntYr2 == "Poor" ~ LambdaYr2 * clim_coef[3],
      AsmntYr2 == "Unsatisfactory" ~ LambdaYr2 * clim_coef[4]
      )) |>
    mutate(ProbYr2 = 1-exp(-LambdaYr2)) |>
    mutate(ExpectedLossYr2 = TotalLoss * ProbYr2) |>
    mutate(RepCostYr2 = ifelse(RepYr2 == 1, ExpectedRepair, 0)) |>
    # NOTE: Savings track cumulatively
    mutate(SavedYr2 = ifelse(RepYr2 >= 1, ExpectedSavingsYrly, 0)) |>
  
    mutate(RepYr3 = ifelse(ExpectedSavings >= saving_splits[[3]], RepYr2 + 1, 0)) |>
    mutate(AsmntYr3 = if_else(RepYr3 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr3 = if_else(RepYr3 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr3 = case_when(
      AsmntYr3 == "Satisfactory" ~ LambdaYr3 * clim_coef[1],
      AsmntYr3 == "Fair" ~ LambdaYr3 * clim_coef[2],
      AsmntYr3 == "Poor" ~ LambdaYr3 * clim_coef[3],
      AsmntYr3 == "Unsatisfactory" ~ LambdaYr3 * clim_coef[4]
      )) |>
    mutate(ProbYr3 = 1-exp(-LambdaYr3)) |>
    mutate(ExpectedLossYr3 = TotalLoss * ProbYr3) |>
    mutate(RepCostYr3 = ifelse(RepYr3 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr3 = ifelse(RepYr3 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr4 = ifelse(ExpectedSavings >= saving_splits[[4]], RepYr3 + 1, 0)) |>
    mutate(AsmntYr4 = if_else(RepYr4 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr4 = if_else(RepYr4 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr4 = case_when(
      AsmntYr4 == "Satisfactory" ~ LambdaYr4 * clim_coef[1],
      AsmntYr4 == "Fair" ~ LambdaYr4 * clim_coef[2],
      AsmntYr4 == "Poor" ~ LambdaYr4 * clim_coef[3],
      AsmntYr4 == "Unsatisfactory" ~ LambdaYr4 * clim_coef[4]
      )) |>
    # mutate(ProbYr4 = ifelse(RepYr4 == 1, NewProbYr, 1-exp(-LambdaYr4))) |>
    mutate(ProbYr4 = 1-exp(-LambdaYr4)) |>
    mutate(ExpectedLossYr4 = TotalLoss * ProbYr4) |>
    mutate(RepCostYr4 = ifelse(RepYr4 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr4 = ifelse(RepYr4 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr5 = ifelse(ExpectedSavings >= saving_splits[[5]], RepYr4 + 1, 0)) |>
    mutate(AsmntYr5 = if_else(RepYr5 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr5 = if_else(RepYr5 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr5 = case_when(
      AsmntYr5 == "Satisfactory" ~ LambdaYr5 * clim_coef[1],
      AsmntYr5 == "Fair" ~ LambdaYr5 * clim_coef[2],
      AsmntYr5 == "Poor" ~ LambdaYr5 * clim_coef[3],
      AsmntYr5 == "Unsatisfactory" ~ LambdaYr5 * clim_coef[4]
      )) |>
    mutate(ProbYr5 = 1-exp(-LambdaYr5)) |>
    mutate(ExpectedLossYr5 = TotalLoss * ProbYr5) |>
    mutate(RepCostYr5 = ifelse(RepYr5 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr5 = ifelse(RepYr5 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr6 = ifelse(ExpectedSavings >= saving_splits[[6]], RepYr5 + 1, 0)) |>
    mutate(AsmntYr6 = if_else(RepYr6 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr6 = if_else(RepYr6 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr6 = case_when(
      AsmntYr6 == "Satisfactory" ~ LambdaYr6 * clim_coef[1],
      AsmntYr6 == "Fair" ~ LambdaYr6 * clim_coef[2],
      AsmntYr6 == "Poor" ~ LambdaYr6 * clim_coef[3],
      AsmntYr6 == "Unsatisfactory" ~ LambdaYr6 * clim_coef[4]
      )) |>
    mutate(ProbYr6 = 1-exp(-LambdaYr6)) |>
    mutate(ExpectedLossYr6 = TotalLoss * ProbYr6) |>
    mutate(RepCostYr6 = ifelse(RepYr6 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr6 = ifelse(RepYr6 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr7 = ifelse(ExpectedSavings >= saving_splits[[7]], RepYr6 + 1, 0)) |>
    mutate(AsmntYr7 = if_else(RepYr7 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr7 = if_else(RepYr7 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr7 = case_when(
      AsmntYr7 == "Satisfactory" ~ LambdaYr7 * clim_coef[1],
      AsmntYr7 == "Fair" ~ LambdaYr7 * clim_coef[2],
      AsmntYr7 == "Poor" ~ LambdaYr7 * clim_coef[3],
      AsmntYr7 == "Unsatisfactory" ~ LambdaYr7 * clim_coef[4]
      )) |>
    mutate(ProbYr7 = 1-exp(-LambdaYr7)) |>
    mutate(ExpectedLossYr7 = TotalLoss * ProbYr7) |>
    mutate(RepCostYr7 = ifelse(RepYr7 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr7 = ifelse(RepYr7 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr8 = ifelse(ExpectedSavings >= saving_splits[[8]], RepYr7 + 1, 0)) |>
    mutate(AsmntYr8 = if_else(RepYr8 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr8 = if_else(RepYr8 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr8 = case_when(
      AsmntYr8 == "Satisfactory" ~ LambdaYr8 * clim_coef[1],
      AsmntYr8 == "Fair" ~ LambdaYr8 * clim_coef[2],
      AsmntYr8 == "Poor" ~ LambdaYr8 * clim_coef[3],
      AsmntYr8 == "Unsatisfactory" ~ LambdaYr8 * clim_coef[4]
      )) |>
    mutate(ProbYr8 = 1-exp(-LambdaYr8)) |>
    mutate(ExpectedLossYr8 = TotalLoss * ProbYr8) |>
    mutate(RepCostYr8 = ifelse(RepYr8 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr8 = ifelse(RepYr8 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr9 = ifelse(ExpectedSavings >= saving_splits[[9]], RepYr8 + 1, 0)) |>
    mutate(AsmntYr9 = if_else(RepYr9 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr9 = if_else(RepYr9 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr9 = case_when(
      AsmntYr9 == "Satisfactory" ~ LambdaYr9 * clim_coef[1],
      AsmntYr9 == "Fair" ~ LambdaYr9 * clim_coef[2],
      AsmntYr9 == "Poor" ~ LambdaYr9 * clim_coef[3],
      AsmntYr9 == "Unsatisfactory" ~ LambdaYr9 * clim_coef[4]
      )) |>
    mutate(ProbYr9 = 1-exp(-LambdaYr9)) |>
    mutate(ExpectedLossYr9 = TotalLoss * ProbYr9) |>
    mutate(RepCostYr9 = ifelse(RepYr9 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr9 = ifelse(RepYr9 >= 1, ExpectedSavingsYrly, 0)) |>
    
    mutate(RepYr10 = ifelse(ExpectedSavings >= saving_splits[[10]], RepYr9 + 1, 0)) |>
    mutate(AsmntYr10 = if_else(RepYr10 == 1, 
                                factor(4, levels=c(1,2,3,4), labels=c("Unsatisfactory","Poor","Fair","Satisfactory")), 
                                NewAssessment, missing=NewAssessment)) |>
    mutate(LambdaYr10 = if_else(RepYr10 == 1, pmin(Lambda, NewLambda), Lambda, missing=Lambda)) |>
    mutate(LambdaYr10 = case_when(
      AsmntYr10 == "Satisfactory" ~ LambdaYr10 * clim_coef[1],
      AsmntYr10 == "Fair" ~ LambdaYr10 * clim_coef[2],
      AsmntYr10 == "Poor" ~ LambdaYr10 * clim_coef[3],
      AsmntYr10 == "Unsatisfactory" ~ LambdaYr10 * clim_coef[4]
      )) |>
    mutate(ProbYr10 = 1-exp(-LambdaYr10)) |>
    mutate(ExpectedLossYr10 = TotalLoss * ProbYr10) |>
    mutate(RepCostYr10 = ifelse(RepYr10 == 1, ExpectedRepair, 0)) |>
    mutate(SavedYr10 = ifelse(RepYr10 >= 1, ExpectedSavingsYrly, 0))
}
```


```{r}
# TODO: Need to also model expected loss changes for ALL dams,
# and repair/savings costs for repaired dams only
# Keep in mind damexps only includes dams that require repairs

# TODO: Expected savings is not adjusted for new probability of failure
# Adjust lambda for each year? ProbYr2 needs to be >=1 for clim_coef[1]==1 or
# need to adjust lambda for repaired dams
# ^^^ Not going to do this, which makes expected savings a conservative estimate
# Make sure to note this assumption in writing

# TODO: plot bond interest to show ability to pay it off?

# Satisfactory, Fair, Poor, Unsatisfactory
clim_coef <- c(1, 1.001, 1.005, 1.01)
# clim_coef <- c(1, 1.0005, 1.0025, 1.005)
# clim_coef <- c(1, 1.0015, 1.01, 1.015)

yearly_inflation <- 1.03

dam_model_climate_low <- model_dams(damexps, c(1, 1.0005, 1.0025, 1.005))
dam_model_climate_mid <- model_dams(damexps, c(1,1.001, 1.005, 1.01))
dam_model_climate_high <- model_dams(damexps, c(1,1.002, 1.01, 1.02))



dam_model_climate_mid |>
  group_by(Region) |>
  filter(RepYr10 >= 1) |>
  summarize(Count = n(),
            RepCost = sum(ExpectedRepair))

damcosts |> group_by(Region) |> summarize(L = sum(ExpectedLoss), LProp = sum(ExpectedLoss) / n())


summary(impeddam)
unique(damdata$Assessment)
```



```{r}
# Calculate variance for ELYi, Si, Ri
dam_model_climate_mid |> 
  mutate(VarLossYr1 = TotalLoss^2 * ProbYr1) |>
  mutate(VarLossYr2 = TotalLoss^2 * ProbYr2) |>
  mutate(VarLossYr3 = TotalLoss^2 * ProbYr3) |>
  mutate(VarLossYr4 = TotalLoss^2 * ProbYr4) |>
  mutate(VarLossYr5 = TotalLoss^2 * ProbYr5) |>
  mutate(VarLossYr6 = TotalLoss^2 * ProbYr6) |>
  mutate(VarLossYr7 = TotalLoss^2 * ProbYr7) |>
  mutate(VarLossYr8 = TotalLoss^2 * ProbYr8) |>
  mutate(VarLossYr9 = TotalLoss^2 * ProbYr9) |>
  mutate(VarLossYr10 = TotalLoss^2 * ProbYr10) |>
  summarize(SD1 = sqrt(sum(VarLossYr1)),
            VL2 = sqrt(sum(VarLossYr2)),
            VL3 = sqrt(sum(VarLossYr3)),
            VL4 = sqrt(sum(VarLossYr4)),
            VL5 = sqrt(sum(VarLossYr5)),
            VL6 = sqrt(sum(VarLossYr6)),
            VL7 = sqrt(sum(VarLossYr7)),
            VL8 = sqrt(sum(VarLossYr8)),
            VL9 = sqrt(sum(VarLossYr9)),
            VL10 = sqrt(sum(VarLossYr10)))

dam_model_climate_mid |> 
  summarize(EL1 = sum(ExpectedLossYr1))

77487.18 + 2 * 7578.154
```

```{r}
dam_end_amt <- function(dam_model, inflation, interest) {
  dam_summary <- dam_model |> 
    summarize(EL1 = sum(ExpectedLossYr1, na.rm=T),
              S1 = sum(SavedYr1, na.rm=T),
              R1 = sum(RepCostYr1, na.rm=T),
              EL2 = sum(ExpectedLossYr2, na.rm=T),
              S2 = sum(SavedYr2, na.rm=T),
              R2 = sum(RepCostYr2, na.rm=T),
              EL3 = sum(ExpectedLossYr3, na.rm=T),
              S3 = sum(SavedYr3, na.rm=T),
              R3 = sum(RepCostYr3, na.rm=T),
              EL4 = sum(ExpectedLossYr4, na.rm=T),
              S4 = sum(SavedYr4, na.rm=T),
              R4 = sum(RepCostYr4, na.rm=T),
              EL5 = sum(ExpectedLossYr5, na.rm=T),
              S5 = sum(SavedYr5, na.rm=T),
              R5 = sum(RepCostYr5, na.rm=T),
              EL6 = sum(ExpectedLossYr6, na.rm=T),
              S6 = sum(SavedYr6, na.rm=T),
              R6 = sum(RepCostYr6, na.rm=T),
              EL7 = sum(ExpectedLossYr7, na.rm=T),
              S7 = sum(SavedYr7, na.rm=T),
              R7 = sum(RepCostYr7, na.rm=T),
              EL8 = sum(ExpectedLossYr8, na.rm=T),
              S8 = sum(SavedYr8, na.rm=T),
              R8 = sum(RepCostYr8, na.rm=T),
              EL9 = sum(ExpectedLossYr9, na.rm=T),
              S9 = sum(SavedYr9, na.rm=T),
              R9 = sum(RepCostYr9, na.rm=T),
              EL10 = sum(ExpectedLossYr10, na.rm=T),
              S10 = sum(SavedYr10, na.rm=T),
              R10 = sum(RepCostYr10, na.rm=T)) |>
    pivot_longer(
      cols=everything(),
      names_to=c("Type", "Year"),
      names_pattern="([S|EL|R])([0-9]{1,2})",
      values_to="Amount"
    ) |>
    mutate(Year = as.numeric(Year)) |> 
    pivot_wider(
      names_from=Type,
      values_from=Amount
    ) |>
    rename(EL = L) |>
    mutate(R = R * inflation ^ (Year - 1),
           S = S * inflation ^ (Year - 1),
           EL = EL * inflation ^ (Year - 1),
           TotalSavings = S * interest ^ Year)
  # Untested annuity
  # |>
  #   mutate(IntS = S * interest ^ seq(9:0))
  # TODO: TotalSavings NOT CUMULATIVELY SUMMED,
  # ALSO DOES NOT SUBTRACT REPAIR COSTS OR BOND SPENDING
  return(dam_summary)
}

# Rates:
# 10yr inflation max avg is 9.2847%, min avg is 1.5671%
# 10yr interest max is 17.41%, min is .97%, avg is 6.71%
# 1yr avg is 4.3%

# Inflation: 3% expected (because US), 
yravg = c(NA)
for(i in seq(1:(nrow(econdata)-9))) {
  sums = c(NA)
  for(j in 1:10) {
    sums[j] = 1 + econdata$Inflation[[i + j - 1]]
  }
  
  yravg[i] = prod(sums) ^ (1/10)
}
infmin <- min(yravg)
infmax <- max(yravg)


yravg = c(NA)
for(i in seq(1:(nrow(econdata)-9))) {
  sums = c(NA)
  for(j in 1:10) {
    sums[j] = 1 + econdata$Spot1Yr[[i + j - 1]]
  }
  
  yravg[i] = prod(sums) ^ (1/10)
}
intmin <- min(yravg)
intavg <- mean(yravg)
intmax <- max(yravg)

# inflationrates <- as.numeric(1 + (econdata |> 
#   summarize(mininf = min(Inflation), 
#             avginf = mean(Inflation),
#             maxinf = max(Inflation)))[1,])
# Mid could be us 10-yr avg, 2.212306%
inflationrates <- c(infmin, 1.03, infmax)

oneyrrates <- as.numeric((econdata |> 
  summarize(min1yr = intmin,
            avg1yr = intavg,
            max1yr = intmax))[1,])

dam_summary_low_inf <- dam_end_amt(dam_model_climate_mid, 
                                   inflationrates[1], oneyrrates[2])
dam_summary_mid_inf <- dam_end_amt(dam_model_climate_mid, 
                                   inflationrates[2], oneyrrates[2])
dam_summary_high_inf <- dam_end_amt(dam_model_climate_mid, 
                                    inflationrates[3], oneyrrates[2])

dam_summary_low_int <- dam_end_amt(dam_model_climate_mid, 
                                   inflationrates[2], oneyrrates[1])
dam_summary_mid_int <- dam_end_amt(dam_model_climate_mid, 
                                   inflationrates[2], oneyrrates[2])
dam_summary_high_int <- dam_end_amt(dam_model_climate_mid, 
                                    inflationrates[2],  oneyrrates[3])

# TODO:
# Will there be enough savings to pay off bond at end?
# How much can acceptably be reserved 

dam_end_amt(dam_model_climate_low, inflationrates[2], oneyrrates[2])

# TODO: For sensitivity analysis, need to adjust inflation so every number is in 
# 2035(4?) Qalkoons
interest = oneyrrates[2]
dam_summary_mid_inf |> mutate(IntS = S * interest ^ rev(seq(0:9))) |>
  summarize(TS = sum(IntS))

# Maybe moving average to get 2025 10-yr spot rate?


impeddam
```

```{r}
# Bond amount
dam_summary_mid_inf |> 
  mutate(IntR = R / interest ^ (Year - 1)) |>
  summarize(TR = sum(IntR))
  # mutate(TR = TR * 1.0103) # TODO: Underwriting fee?

library(forecast)
ma_10yr <- auto.arima(econdata$Spot10Yr)
ma_10yr
ts.plot(econdata$Spot10Yr)
tenyrint <- forecast(ma_10yr)$mean[1]
tenyrint

econdata[63, "Spot10Yr"]

ma_1yr <- auto.arima(econdata$Spot1Yr)
oneyrint <- forecast(ma_1yr)$mean[1]
oneyrint

econdata[63, "Spot1Yr"]
```


Bond amount – Program setup cost (underwriting cost + setup cost + labor cost) + repair costs for all ten(?) years, 10 year interest rate for redemption value
Analysis – Inflow pays for inspection costs, expected repair, 95% interval on expected loss including climate change; expected savings to pay off bond



Bond notes:
given country's wealth and similarity to US, assume currency stable and can issue bonds in our currency
bonds in own country, compare to own country rates; overseas compare to overseas rates
selling bond in own country backed by legislature; risk-free and can use rates
assume bonds backed by full government, risk-free, country big enough to have robust bond
assume paying out 10yr, investment yield 1yr



```{r}
damsums <- dam_model_climate_mid |>
  summarize(EL1 = sum(ExpectedLossYr1),
            S1 = sum(SavedYr1),
            R1 = sum(RepCostYr1),
            EL2 = sum(ExpectedLossYr2),
            S2 = sum(SavedYr2),
            R2 = sum(RepCostYr2),
            EL3 = sum(ExpectedLossYr3),
            S3 = sum(SavedYr3),
            R3 = sum(RepCostYr3),
            EL4 = sum(ExpectedLossYr4),
            S4 = sum(SavedYr4),
            R4 = sum(RepCostYr4),
            EL5 = sum(ExpectedLossYr5),
            S5 = sum(SavedYr5),
            R5 = sum(RepCostYr5),
            EL6 = sum(ExpectedLossYr6),
            S6 = sum(SavedYr6),
            R6 = sum(RepCostYr6),
            EL7 = sum(ExpectedLossYr7),
            S7 = sum(SavedYr7),
            R7 = sum(RepCostYr7),
            EL8 = sum(ExpectedLossYr8),
            S8 = sum(SavedYr8),
            R8 = sum(RepCostYr8),
            EL9 = sum(ExpectedLossYr9),
            S9 = sum(SavedYr9),
            R9 = sum(RepCostYr9),
            EL10 = sum(ExpectedLossYr10),
            S10 = sum(SavedYr10),
            R10 = sum(RepCostYr10)) 

damsumsyrly <- damsums |>
  pivot_longer(
    cols=everything(),
    names_to=c("Type", "Year"),
    names_pattern="([S|EL|R])([0-9]{1,2})",
    values_to="Amount"
  ) |>
  mutate(Year = as.numeric(Year)) |>
```

```{r}
# Nice looking table, bad for graphing
damsumsyrly |> 
  pivot_wider(
    names_from=Type,
    values_from=Amount
  ) |>
  rename(EL = L)


dam_model_climate_mid |> 
    summarize(EL1 = sum(ExpectedLossYr1, na.rm=T),
              S1 = sum(SavedYr1, na.rm=T),
              R1 = sum(RepCostYr1, na.rm=T),
              EL2 = sum(ExpectedLossYr2, na.rm=T),
              S2 = sum(SavedYr2, na.rm=T),
              R2 = sum(RepCostYr2, na.rm=T),
              EL3 = sum(ExpectedLossYr3, na.rm=T),
              S3 = sum(SavedYr3, na.rm=T),
              R3 = sum(RepCostYr3, na.rm=T),
              EL4 = sum(ExpectedLossYr4, na.rm=T),
              S4 = sum(SavedYr4, na.rm=T),
              R4 = sum(RepCostYr4, na.rm=T),
              EL5 = sum(ExpectedLossYr5, na.rm=T),
              S5 = sum(SavedYr5, na.rm=T),
              R5 = sum(RepCostYr5, na.rm=T),
              EL6 = sum(ExpectedLossYr6, na.rm=T),
              S6 = sum(SavedYr6, na.rm=T),
              R6 = sum(RepCostYr6, na.rm=T),
              EL7 = sum(ExpectedLossYr7, na.rm=T),
              S7 = sum(SavedYr7, na.rm=T),
              R7 = sum(RepCostYr7, na.rm=T),
              EL8 = sum(ExpectedLossYr8, na.rm=T),
              S8 = sum(SavedYr8, na.rm=T),
              R8 = sum(RepCostYr8, na.rm=T),
              EL9 = sum(ExpectedLossYr9, na.rm=T),
              S9 = sum(SavedYr9, na.rm=T),
              R9 = sum(RepCostYr9, na.rm=T),
              EL10 = sum(ExpectedLossYr10, na.rm=T),
              S10 = sum(SavedYr10, na.rm=T),
              R10 = sum(RepCostYr10, na.rm=T)) |>
    pivot_longer(
      cols=everything(),
      names_to=c("Type", "Year"),
      names_pattern="([S|EL|R])([0-9]{1,2})",
      values_to="Amount"
    ) |>
    # mutate(R = R / inflation ^ (Year - 1),
      #      S = S / inflation ^ Year,
      #      EL = EL / inflation ^ (Year - 1),
      #      IntS = S * interest ^ seq(9:0)) |> 
  mutate(Year = as.numeric(Year)) |>
  ggplot(aes(x=Year, y=Amount, color=Type)) + 
    geom_point() +
    geom_line() +
    labs(title="Expected Loss and Savings over time",
         subtitle = "Data in millions of Qalkoons") +
    scale_x_continuous(breaks=seq(1,10), labels=seq(1,10)) + 
    scale_y_continuous(breaks=seq(0,1000000,5000)) + 
    scale_color_discrete(labels=c("Loss", "Repair", "Savings"))
    
ggsave("DamLossSavingsTimeFull.png", width=1920, height=1080, units="px")
```

